---
title: "Capiral_Margaret_Lab2"
output: html_document
date: "2026-08-24"
---
Bio 118 Laboratory Exercise 2

Name: Margaret Capiral

Section: A2

Date: August 24, 2026 

GitHub repository URL: https://github.com/margaretcapiral/BIO118.git

# Introduction

R uses different data types and structures to organize and work with information in data analysis. In this activity, the basic data types and operators in R are introduced through biological and environmental examples. The activity also focuses on common data structures, including vectors, matrices, data frames, lists, and factors. Understanding these concepts is important for properly storing, accessing, and manipulating biological data.

# Laboratory Exercise 2A

## 2.3 Practice

### 2.3.1 Monitoring Species Population in Different Habitats

```{r}
species_name <- "Setophaga petechia"
species_upper <- toupper(species_name)
print(species_upper) 
class(species_upper) 

habitat_A_population <- 320
habitat_B_population <- 275

total_population <- habitat_A_population + habitat_B_population
print (total_population)

is_habitat_A_larger <- habitat_A_population > habitat_B_population
print (is_habitat_A_larger)

population_difference <- habitat_A_population - habitat_B_population
print(population_difference)
```

### 2.3.2 Analyzing Environmental Data for Climate Change Impact

```{r}
current_temperature <- 25
print (current_temperature)
predicted_increase <- 3.8
print (predicted_increase)
future_temperature <- current_temperature + predicted_increase
print (future_temperature)
current_temperature_farenheit <- (current_temperature * 9/5) + 32
print (current_temperature_farenheit)
future_temperature_farenheit <- (future_temperature * 9/5) + 32
print (future_temperature_farenheit)
exceeds_threshold <- future_temperature > 5
print (exceeds_threshold)
```

### 2.3.3 Evaluating Pollutant Levels in a River

```{r}
chemical_X_concentration <- 15.3
print (chemical_X_concentration)
chemical_Y_concentration <- 9.8
print (chemical_Y_concentration)
is_chemical_X_safe <- chemical_X_concentration < 10
print (is_chemical_X_safe)
is_chemical_Y_safe <- chemical_Y_concentration < 10
print (is_chemical_Y_safe)
are_both_safe <- is_chemical_X_safe & is_chemical_Y_safe
print (are_both_safe)
```

# Laboratory Exercise 2B: Understanding Data Structures

## 2.2 Practice

### 2.2.1 Working with DNA Sequences

```{r}
dna_sequences <- c("ATGCGTACGTAGCTAGCGT", "ATGCGTACGTAGTTAGCGT", "ATGCGTACGTAGATAGCGT")
nchar(dna_sequences)
library(stringr)
str_count(dna_sequences, "A")
```

### 2.2.2 Monitoring Species Population in Different Habitats
```{r}
biodiversity_data <- data.frame(
  Species = c("Oak", "Pine", "Maple", "Birch", "Spruce", 
              "Cedar", "Ash", "Fir", "Beech"),
  Count = c(120, 85, 100, 60, 150, 75, 55, 95, 110),
  Habitat = c("Forest_A", "Forest_A", "Forest_A",
              "Forest_B", "Forest_B", "Forest_B",
              "Forest_C", "Forest_C", "Forest_C")
)

print(biodiversity_data)

total_per_habitat <- tapply(
  biodiversity_data$Count,
  biodiversity_data$Habitat,
  sum
)

print(total_per_habitat)

species_less_than_60 <- subset(biodiversity_data, Count < 60)

print(species_less_than_60)
```

### 2.2.3 Monitoring Temperature Changes
```{r}
monthly_temps <- matrix(
  c(2.5, 3.1, 6.2, 10.5, 15.3, 20.1, 22.4, 21.9, 18.3, 13.8, 7.5, 3.9,
    2.7, 3.5, 6.8, 11.0, 16.0, 21.0, 23.1, 22.5, 19.0, 14.5, 8.0, 4.1),
  nrow = 12,
  ncol = 2
)

rownames(monthly_temps) <- c(
  "January", "February", "March", "April", "May", "June",
  "July", "August", "September", "October", "November", "December"
)

colnames(monthly_temps) <- c("Year_1", "Year_2")

print(monthly_temps)

average_temps <- colMeans(monthly_temps)

print(average_temps)

year_2_higher <- average_temps["Year_2"] > average_temps["Year_1"]

print(year_2_higher)

summer_temps <- monthly_temps[c("June", "July", "August"), ]

print(summer_temps)
```

# Reflection
This activity helped me better understand how different data types and structures are used in R. I learned how operators can be applied to biological and environmental data to perform calculations and make comparisons. Working with vectors, matrices, data frames, lists, and factors also helped me see how different types of information can be organized. This laboratory activity made the basic concepts of working with data in R clearer and more practical.

# Conclusion
In conclusion, the activities focused on understanding data types, operators, and data structures in R. We practiced creating and manipulating different types of data and using arithmetic, relational, and logical operators. We also worked with vectors, matrices, data frames, lists, and factors to organize and retrieve biological data. These activities provided a foundation for handling and analyzing biological datasets using R.